---
title: "Conditioning, follow-up, and at-risk masks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditioning, follow-up, and at-risk masks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette focuses on *conditional* forecasting: summarizing outcomes among runs that meet a definition such as "still under active follow-up" or "event-free through a given time".

## Two core flags: `alive` and `active_followup`

In this ecosystem, we keep two concepts separate:

- `alive`: vital status only (TRUE or FALSE).
- `active_followup`: whether the patient is still being followed / in scope for the question.

A patient can be discharged (or administratively censored) while still alive: `alive == TRUE` but `active_followup == FALSE`.

These are **state variables**. Model code should update them explicitly inside `transition()`.

## Setup

```{r}
library(patientSimCore)
library(patientSimASCVD)
library(patientSimForecast)

bundle <- patientSimASCVD::bundle_ascvd()
engine <- patientSimCore::Engine$new(bundle = bundle)
```

## Conditioning at a start time (`eligible`)

All streaming summaries in patientSimForecast define a *fixed cohort* at a `start_time` (default: the first forecast time). By default, the eligible cohort is:

- defined at `start_time`, and
- `alive == TRUE` at `start_time`.

You can add additional conditions by supplying an `eligible()` function. This function is evaluated **once** at `start_time`.

```{r}
p <- patientSimCore::new_patient(
  schema = patientSimASCVD::ascvd_schema(),
  state = list(
    age = 55,
    sex = "M",
    sbp = 148,
    dbp = 92,
    ldl = 165,
    hdl = 42,
    triglycerides = 210,
    n_antihypertensives = 0,
    statin_intensity = "none",
    alive = TRUE,
    active_followup = TRUE
  ),
  time0 = 0
)

times <- c(1/12, 3/12, 6/12, 1, 2)
start_time <- 3/12

eligible_active_followup <- function(snap, t, ctx) {
  isTRUE(snap$active_followup)
}

rf <- patientSimForecast::forecast(
  engine = engine,
  patients = list(p = p),
  times = times,
  S = 200,
  return = "summary_stats",
  summary_stats = "risk",
  summary_spec = list(
    event = c("mi"),
    start_time = start_time,
    eligible = eligible_active_followup
  )
)
rf$result
```

## Event-free conditioning through the start time

Sometimes you want "risk over the next year among people who have had no event so far".

Use `condition_on_events` to require that a set of event types has **not** occurred by `start_time`.

```{r eval=FALSE}
rf_event_free <- patientSimForecast::forecast(
  engine = engine,
  patients = list(p = p),
  times = times,
  S = 200,
  return = "summary_stats",
  summary_stats = "risk",
  summary_spec = list(
    event = c("mi"),
    start_time = start_time,
    condition_on_events = c("mi")
  )
)
```

## A note on "at-risk" masks

Many real models need a time-varying "at risk" definition, such as "alive and still on the waitlist" (transplant) or "alive and currently on therapy" (oncology).

In v1, patientSimForecast uses a fixed cohort defined at `start_time`. You can still summarize *state variables* at future times, but the meaning of the denominator matters.

Practical guidance:

- Use `eligible()` to define a clinically meaningful cohort at a clear anchor time.
- Prefer defining additional phase/status variables in state (or as derived variables), so you can condition on them cleanly.
- When you need a true time-varying at-risk denominator (re-entry allowed), we can extend the API with an `at_risk()` predicate.
