---
title: "Forecasting with patientSimForecast (ASCVD example)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting with patientSimForecast (ASCVD example)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What forecasting means here

In this ecosystem, a *forecast* means:

- Start from a patient's current state (age, sex, labs, treatments, etc.).
- Clone that patient many times.
- Run the model forward for each clone.
- Summarize the distribution of outcomes across clones at a set of future times.

The result is a probabilistic forecast, not a single deterministic trajectory.

## Setup

```{r}
library(patientSimCore)
library(patientSimASCVD)
library(patientSimForecast)
```

We will use the ASCVD example model bundle and engine.

```{r}
bundle <- patientSimASCVD::bundle_ascvd()
engine <- patientSimCore::Engine$new(bundle = bundle)
```

## Create a "real" patient by injecting state

The easiest way to forecast for a real patient is to create a `Patient` and provide the core state variables directly.

```{r}
# 55 year-old male with elevated BP and lipids
p <- patientSimCore::new_patient(
  schema = patientSimASCVD::ascvd_schema(),
  state = list(
    age = 55,
    sex = "M",
    sbp = 148,
    dbp = 92,
    ldl = 165,
    hdl = 42,
    triglycerides = 210,
    n_antihypertensives = 0,
    statin_intensity = "none",
    alive = TRUE,
    active_followup = TRUE
  ),
  time0 = 0
)
```

## Choose forecast times

The ASCVD example uses **years** as the time unit.

```{r}
times <- c(1/12, 3/12, 6/12, 1, 2)  # 1, 3, 6 months; 1 and 2 years
```

## Risk of an event

A memory-light workflow is to compute risk directly, without keeping all simulated trajectories.

```{r}
# Risk of composite ASCVD events (as defined by the model)
rf <- patientSimForecast::forecast(
  engine = engine,
  patients = list(p = p),
  times = times,
  S = 2000,
  return = "summary_stats",
  summary_stats = "risk",
  summary_spec = list(event = c("ascvd_event")),
  backend = "future",
  seed = 1
)
rf$result
```

## State summaries

You can also summarize future state variables (numeric and categorical).

```{r}
ss <- patientSimForecast::forecast(
  engine = engine,
  patients = list(p = p),
  times = times,
  S = 200,
  return = "summary_stats",
  summary_stats = "state",
  summary_spec = list(vars = c("sbp", "dbp", "ldl", "n_antihypertensives", "statin_intensity")),
  backend = "none",
  seed = 1
)

# Numeric summaries are returned as one table per variable
ss$numeric$sbp

# Categorical summaries are returned as one table per variable
ss$categorical$statin_intensity
```

## Parallel runs

For large jobs (many patients or many simulations), use a parallel backend.

- `backend = "mclapply"` uses `parallel::mclapply()` (macOS/Linux only).
- `backend = "future"` uses `future.apply::future_lapply()` (recommended for clusters/cloud).

With `backend = "future"`, you typically set your plan once:

```{r eval=FALSE}
library(future)
plan(multisession, workers = 4)

rf <- patientSimForecast::forecast(
  engine = engine,
  patients = list(p = p),
  times = times,
  S = 2000,
  return = "summary_stats",
  summary_stats = "risk",
  summary_spec = list(event = c("ascvd_event")),
  backend = "future",
  seed = 1
)
```
